---
title: "Annotation"
author: "NadineBestard"
date: "30/03/2021"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
library(here) #reproducible paths
library(scater) #feature plots
library(patchwork) # agregate plots
library(scran) # for findmarkers
```

```{r age}
project <- "fire-mice"
```

```{r import}
if (!file.exists(here("processed", project, "sce_anno_01.RDS"))) {
  sce <- readRDS(here("processed", project, "sce_clusters_01.RDS"))
}else{
  sce <- readRDS(here("processed", project, "sce_anno_01.RDS"))
}
```
## Clustering at k40
```{r}
plotTSNE(sce, colour_by = "cluster_k40", text_by = "cluster_k40") + scale_color_hue()
```

## Annotation

We make a first rough annotation using known markers for different
celltypes.

<details>
  <summary>Click to expand the Neurons marker plots</summary>
```{r 4-feature-plots, cache=TRUE}
#Neurons:
list_plots <- lapply(c("Snap25", "Stmn2", "Rbfox3", "Gabrb2"),
                     function(x)plotTSNE(sce, colour_by = x ))

wrap_plots(list_plots) +  plot_annotation(title = "Neurons")
```

```{r violin-neuro, fig.height=4}
plotExpression(sce, features=c("Snap25", "Stmn2", "Rbfox3", "Gabrb2") ,
    x="cluster_k40", colour_by = "cluster_k40", ncol=1) + scale_color_hue()
```

```{r 4-feature-plots-default, cache=TRUE}
#Inhibitory neurons

list_plots <- lapply(c("Gad1", "Gad2", "Slc32a1", "Pvalb"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  "Inhibitory Neurons")
```

```{r violin-neuroin, fig.height=4, cache=TRUE}
plotExpression(sce, features=c("Gad1", "Gad2", "Slc32a1", "Pvalb"),
    x="cluster_k40", colour_by = "cluster_k40", ncol=1) + scale_color_hue()
```

```{r one-line, fig.height=3, fig.width=9, cache=TRUE}
# Excitatory neurons

list_plots <- lapply(c("Satb2", "Slc12a6", "Slc17a7"),
                     function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title =  "Exitatory Neurons")
```

```{r violin-neuroex, fig.height=3, cache=TRUE}
plotExpression(sce, features=c("Satb2", "Slc12a6", "Slc17a7"),
    x="cluster_k40", colour_by = "cluster_k40", ncol=1) + scale_color_hue()
```

```{r 4-plots, cache=TRUE}
# RBFOX3 (and other Granulate cell markers)
list_plots <- lapply(c("Cdh15", "Calb2", "Rbfox3", "Reln"),
function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title =  "RBFO3+ Neurons")
```

```{r violin-neurorbfox, fig.height=4, cache=TRUE}
plotExpression(sce, features=c("Cdh15", "Calb2", "Rbfox3", "Reln"),
    x="cluster_k40", colour_by = "cluster_k40", ncol=1) + scale_color_hue()
```

</details>

<details>
  <summary>Click to expand the Stromal  marker plots</summary>
```{r small-feature-plots2, cache=TRUE}
#Stromal

list_plots <- lapply(c( "Lamb1" , 
                                     "Hspg2", 
                                     "Col4a1", 
                                     "Fn1", 
                                     "Lama2"),
            function(x) plotTSNE(sce, colour_by = x ))

wrap_plots(list_plots) +  plot_annotation(title =  "Stromal")
```

```{r violin-stromal, fig.height=5, cache=TRUE}
plotExpression(sce, features=c( "Lamb1" , 
                                     "Hspg2", 
                                     "Col4a1", 
                                     "Fn1", 
                                     "Lama2"),
    x="cluster_k40", colour_by = "cluster_k40", ncol=1) + scale_color_hue()
```

```{r 3lines-feature-plots, fig.height=8, cache=TRUE}
# Endothelial cells and pericytes

list_plots <- lapply(c( "Cldn5",   
                                     "Icam2",
                                     "Pdgfrb", 
                                     "Notch3", 
                                     "Vwf",
                                     "Flt1",
                                     "Mecom"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  " Endothelial cells and pericytes")
```

```{r violin-endo, fig.height=7, cache=TRUE}
plotExpression(sce, features=c( "Cldn5",   
                                     "Icam2",
                                     "Pdgfrb", 
                                     "Notch3", 
                                     "Vwf",
                                     "Flt1",
                                     "Mecom"),
    x="cluster_k40", colour_by = "cluster_k40", ncol=1) + scale_color_hue()
```

</details>

<details>
  <summary>Click to expand the Epithelial cells</summary>
```{r one-line-3-col, fig.height=4, cache=TRUE}
list_plots <- lapply(c("Ttr", "Kcnj13", "Krt18"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  "Choroid plexus epithelial cells")
```

```{r violin-epi, fig.height=3, cache=TRUE}
plotExpression(sce, features=c("Ttr", "Kcnj13", "Krt18"),
    x="cluster_k40", colour_by = "cluster_k40", ncol=1) + scale_color_hue()
```

</details>

<details>
  <summary>Click to expand the Astrocytes marker plots</summary>

```{r 3lines-4-col-feature-plots, fig.height=7, fig.width=9, cache=TRUE}
# Astrocytes

list_plots <- lapply(c("Gja1",
                                    "Aqp4", 
                                    "Glul", 
                                    "Sox9", 
                                    "Ndrg2", 
                                    "Gfap", 
                                    "Aldh1a1", 
                                    "Aldh1l1", 
                                    "Vim", 
                                    "Apoe", 
                                    "Fgfr3"),
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title = "Astrocyte")
```

```{r violin-astro, fig.height=11, cache=TRUE}
plotExpression(sce, features=c("Gja1",
                                    "Aqp4", 
                                    "Glul", 
                                    "Sox9", 
                                    "Ndrg2", 
                                    "Gfap", 
                                    "Aldh1a1", 
                                    "Aldh1l1", 
                                    "Vim", 
                                    "Apoe", 
                                    "Fgfr3"),
    x="cluster_k40", colour_by = "cluster_k40", ncol=1) + scale_color_hue()
```

```{r 3lines-feature-plots-2, cache=TRUE, fig.height=9, fig.width=9}
#Astrocyte markers as described in Zeisel et al. 2018 in the mouse for telencephalon and non-telencephalon astrocytes

list_plots <- lapply(c( "Agt", 
                                     "Mfge8",  
                                     "Slc6a11",
                                     "Slc6a9", 
                                     "Gdf10",
                                     "Islr",
                                     "Gfap",
                                     "Aqp4")  ,
                     function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title = "Astrocyte mouse telecephalon")
```

</details>

<details>
  <summary>Click to expand the Immune cells plots</summary>

```{r 3lines-feature-plots3, cache=TRUE, fig.height=8}
#Microglia and macrophages

list_plots <- lapply(c( "Cd74", 
                                     "Spi1", 
                                     "Mrc1", 
                                     "Tmem119", 
                                     "Cx3cr1", 
                                     "Aif1",
                                     "P2ry12",
                                     "C1qc",
                                     "C1qa"),
            function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title = "Microglia and macrophages")
```

```{r violin-micro, fig.height=9, cache=TRUE}
plotExpression(sce, features=c( "Cd74", 
                                     "Spi1", 
                                     "Mrc1", 
                                     "Tmem119", 
                                     "Cx3cr1", 
                                     "Aif1",
                                     "P2ry12",
                                     "C1qc",
                                     "C1qa"),
    x="cluster_k40", colour_by = "cluster_k40", ncol=1) + scale_color_hue()
```

```{r  3-in-one-line, fig.height=4, cache=TRUE}
# Border associated mcrophages
list_plots <- lapply(c( "Mrc1", "Ms4a7", "Apoe"),
            function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title = "Border associated macrophages")
```

```{r violin-bam, fig.height=3, cache=TRUE}
plotExpression(sce, features=c( "Mrc1", "Ms4a7", "Apoe"),
    x="cluster_k40", colour_by = "cluster_k40", ncol=1) + scale_color_hue()
```

```{r  two-lines, fig.height=8, cache=TRUE}
# Immune cells 
list_plots <- lapply(c( 
  # Tcells
  "Cd3e", 
  # Bcells
  "Cd19", 
  # Natural killer
  "Klrb1c", "Cd209a", 
  # all immune cells (CD45)
  "Ptprc"),
            function(x)plotTSNE(sce, 
                                colour_by = x ,
                                point_alpha=0.3,
                                point_size = 0.5)) 
wrap_plots(list_plots) +  plot_annotation(title = "Immune cells")
```

```{r violin-immuno, fig.height=5, cache=TRUE}
plotExpression(sce, features=c( 
  # Tcells
  "Cd3e", 
  # Bcells
  "Cd19", 
  # Natural killer
  "Klrb1c", "Cd209a", 
  # all immune cells (CD45)
  "Ptprc"),
    x="cluster_k40", colour_by = "cluster_k40", ncol=1) + scale_color_hue()
```

```{r fig.height=4}
#monocybes/neutrophils
list_plots <- lapply(c( 
  # monocytes
  "S100a9",
  #neutrophils
  "Ly6g", "Camp"),
            function(x)plotTSNE(sce, 
                                colour_by = x ,
                                point_alpha=0.3,
                                point_size = 0.5)) 
wrap_plots(list_plots) +  plot_annotation(title = "Monocytes and Neutorphils cells")
```

```{r violin-mono, fig.height=3, cache=TRUE}
plotExpression(sce, features=c( 
   # monocytes
  "S100a9",
  #neutrophils
  "Ly6g", "Camp"),
    x="cluster_k40", colour_by = "cluster_k40", ncol=1) + scale_color_hue()
```

</details>

<details>
  <summary>Click to expand the Oligodendroglia marker plots</summary>
```{r big-feature-plots4, cache=TRUE, fig.height=8}
#OPCs

list_plots <- lapply(c("Pdgfra", 
                                    "Cspg4", 
                                    "Gpr17", 
                                    "Ptprz1",
                                    "Olig1", 
                                    "Olig2", 
                                    "Pcdh15", 
                                    "Ptgds",
                                    "Bcan"),
            function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title = "OPCS")
```

```{r violin-opc, fig.height=9, cache=TRUE}
plotExpression(sce, features=c("Pdgfra", 
                                    "Cspg4", 
                                    "Gpr17", 
                                    "Ptprz1",
                                    "Olig1", 
                                    "Olig2", 
                                    "Pcdh15", 
                                    "Ptgds",
                                    "Bcan"),
    x="cluster_k40", colour_by = "cluster_k40", ncol=1) + scale_color_hue()
```

```{r big-feature-plots5, cache=TRUE, fig.height=8}
#Oligodendrocytes

list_plots <- lapply(c("Plp1", 
                                    "Cnp", 
                                    "Mag", 
                                    "Mog", 
                                    "Mobp", 
                                    "Mbp", 
                                    "Sox10" ), 
            function(x)plotTSNE(sce, colour_by = x ))
wrap_plots(list_plots) +  plot_annotation(title =  "Oligodendrocytes")
```

```{r violin-oligo, fig.height=7, cache=TRUE}
plotExpression(sce, features=c("Plp1", 
                                    "Cnp", 
                                    "Mag", 
                                    "Mog", 
                                    "Mobp", 
                                    "Mbp", 
                                    "Sox10" ),
    x="cluster_k40", colour_by = "cluster_k40", ncol=1) + scale_color_hue()
```

</details>

<details>
  <summary>Click to expand the ependymal marker plots</summary>
```{r ependymal, fig.height=8, fig.width=9, cache=TRUE}
list_plots <- lapply(c("Vit", "Sox9", "Dynlrb2", "Ccdc153", "Rsph1", "Tm4sf1", "Pcp4l1", "Pcp4", "Hspa2", "Cd24a", "Mt2", "Chchd10"), 
           function(x)plotTSNE(sce, colour_by = x )) 
wrap_plots(list_plots) +  plot_annotation(title =  "Ependymal")

```

</details>

## Generate marker list

We compute here a pairwise differential expression between all the clusters. The results are saved in a list of dataframes, 
one for each cluster. 


```{r}
# compute markers
if(!file.exists(here("outs", project, "markers_k40_01.RDS"))){
  markers <- findMarkers(sce, groups = sce$cluster_k40, direction="up", lfc=1)
  saveRDS(markers, here("outs", project, "markers_k40_01.RDS"))
}else{
  markers <- readRDS(here("outs", project, "markers_k40_01.RDS"))
}
```

## Rename the Clusters with assigned Cell Types

For this first annotation we will rename the levels from the k=40
clustering as follows
1 

2 

3 

4 

5 

6 

7 

8 

9 

10

11

12

13

14

15

16

17

18

19

20

21

```{r result, eval=FALSE, include=FALSE}
if (!file.exists(here("processed", project, "sce_anno_01.RDS"))) {
sce$celltype <- sce$cluster_k40 
# add the celltypes as described above in the correct order to replace the levels.
levels(sce$celltype)<-c("")
 saveRDS(sce, here("processed", project, "sce_anno_01.RDS"))
}
plotTSNE(sce, colour_by = "celltype", text_by = "cluster_k40") + scale_color_hue()

```
