---
title: "Batch correction"
author: "NadineBestard"
date: "08/03/2021"
output: html_document
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set-up

```{r set-up, message=FALSE, warning=FALSE}
library(here) # for reproducible paths
library(SingleCellExperiment)
library(scater) # For qc and visualisation
library(scran) # For normalisation
library(Matrix) # For log transorming the raw data
library(ggplot2) # To add titles to plots
library(batchelor) # Batch correction
```

```{r matlog2}
# Adapted function from VISION to log tranform sparse matrix
# I could not download the package
matLog2 <- function(spmat, scale = FALSE, scaleFactor = 1e6) {


    if (scale == TRUE) {
        spmat <- t( t(spmat) / colSums(spmat)) * scaleFactor
    }

    if (is(spmat, "sparseMatrix")) {
        matsum <- summary(spmat)

        logx <- log2(matsum$x + 1)

        logmat <- sparseMatrix(i = matsum$i, j = matsum$j,
                               x = logx, dims = dim(spmat),
                               dimnames = dimnames(spmat))
    } else {
        logmat <- log2(spmat + 1)
    }


    return(logmat)

}

```

```{r project}
project <- "fire-mice"
```

# Batch correction with MNN

Combat was shown to outperform other batch correction methods for simple batch
correction (Buttner et. al, 2019). However, this will also regress other
biological differences that are not well balanced between batches. Integration
techniques account for this fact, with the downside that it can lead to
over-correction due to increased degrees of freedom of these non-linear methods.

We use a merge tree, useful for merging together batches that are known to be
more closely related (same tissue) before attempting difficult merges involving
more dissimilar batches.

```{r import}

```

```{r batch-correct}
if (!(file.exists(
  here("processed", project,  "sce_corrected.RDS")
))) {
sce <- readRDS(here("processed", project,  "sce_dimred_01.RDS"))
set.seed(100)
sce <- correctExperiments(sce,
  batch = factor(sce$chip),
  subset.row = rowSubset(sce),
  correct.all=TRUE,
  PARAM = FastMnnParam(
  merge.order = 
    list(list("3","5"), list("4","6")),
  d = 25,
  prop.k=0.10
  )
)
} else {
   sce <- readRDS(here("processed", project,  "sce_corrected.RDS"))
}
```

### Assess Confunding factors impact

*For seek of time this is skipped -This plot is not meant to be used with the
new gene values, as these are just a reconstruction that should only be used for
visual purpose. I was still curious to see how it would look like.-*

```{r , eval=F}
# Only compute if first time
if (!(file.exists(here("processed", project,  "variance_explained_corrected.RDS")))) {
  # Log transfrom the corrected values
  assay(sce, "log_reconstructed") <- matLog2(assay(sce, "reconstructed"))
  # Calculate the matrix (time consuming step)
  var_corrected <- getVarianceExplained(
    sce,
    exprs_values = "log_reconstructed",
    variables = c(
      "tissue",
      "chip",
      "age",
      "genotype",
      "mouse",
      "total"
    )
  )
  saveRDS(var_corrected, here("processed", project,  "variance_explained_corrected.RDS"))
  #If not just load created object
} else{
  var_corrected <- readRDS(here("processed", project,  "variance_explained_corrected.RDS"))
}
plotExplanatoryVariables(var_corrected)
```

```{r}
plotReducedDim(sce, colour_by= "chip", point_size=0.1, dimred = "corrected") + 
      ggtitle("After batch correction, small dots")
```

```{r tsne}
if (!(file.exists(
  here("processed", project,  "sce_corrected.RDS")
))) {
#keep the previous dimensional reduction just in case
reducedDim(sce, "TSNE_uncorrected") <- reducedDim(sce, "TSNE")
set.seed(100)
sce <- runTSNE(sce,  dimred="corrected")
}
# plot uit
plotReducedDim(sce, colour_by= "chip", point_size=0.1, dimred = "TSNE_uncorrected") + 
      ggtitle("TSNE dimensional reduction before correction")
plotReducedDim(sce, colour_by= "chip", point_size=0.1, dimred = "TSNE") + 
      ggtitle("TSNE dimensional reduction corrected")
```

One useful diagnostic is the proportion of variance within each batch that is
lost during MNN correction. Specifically, this refers to the within-batch
variance that is removed during orthogonalization with respect to the average
correction vector at each merge step. This is in the metadata, which contains a
matrix of the variance lost in each batch (column) at each merge step (row).

Large proportions of lost variance (\>10%) suggest that correction is removing
genuine biological heterogeneity. This would occur due to violations of the
assumption of orthogonality between the batch effect and the biological
subspace. In this case, the proportion of lost variance is small, indicating
that non-orthogonality is not a major concern.

```{r}
metadata(sce)$merge.info$lost.var
```

```{r}
if (!(file.exists(
  here("processed", project,  "sce_corrected.RDS")
))) {  saveRDS(sce, here("processed", project,  "sce_corrected.RDS"))}
```

## Session Info

<details>

<summary>

Click to expand

</summary>

```{r session-info}
sessionInfo()
```

</details>
