---
title: "Compare DE methods"
author: "NadineBestard"
date: "06/05/2021"
output: html_document
---


I tested a few different methods for DE:

MAST. Listed in [this](https://www.nature.com/articles/nmeth.4612) study as one of the tops methods for sc DE as it adds the detection rate (the fraction of genes expressed in a cell) as a covariate

edgeR. Listed in [this](https://www.nature.com/articles/nmeth.4612) study as one of the top methods for sc DE, it easily allows to add covariates to the design, I added the batch as a covariate. 

edgeR + detection rate: The best method in [this](https://www.nature.com/articles/nmeth.4612)  benchmark study (adding the batch as a covariate but also the detection rate like MAST does)

MAST + batch: I added to the MAST method another covariate to account for the batch as a latent variable. 
https://www.nature.com/articles/nmeth.4612/figures/5 

### Set-up


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# compare results DE

Mast outputs by default the top 120 genes, I will extract the top 120 genes from the DE done with edgeR in order to compare them. 

```{r}
# libraries and setup
library(here) # paths reproducible
library(edgeR) # DE import from edgeR
library(dplyr) #manipulatedf
project <- "fire-mice"
# import
de_results_edgeR <- readRDS(here("processed", project, "DE_oligo_edgeR_de_results.RDS"))
de_results_edgeR_dr <- readRDS(here("processed", project, "DE_oligo_edgeR_detRate_de_results.RDS"))
top_mast <- readRDS(here("processed", project, "DE_oligo_mast_de_results.RDS"))
top_mast_batch <- readRDS(here("processed", project, "DE_oligo_mast_batchcorrect_de_results.RDS"))
top_edge_dr <- data.frame(topTags(de_results_edgeR_dr, n = 120))
top_edge <- data.frame(topTags(de_results_edgeR, n = 120))
```

# Explore each dataset

Results in the order: MAST(dr), edgeR(b), edgeR(b)+dr, MAST(dr)+b
```{r}
tops <- list(top_mast, top_edge, top_edge_dr, top_mast_batch)
names(tops) <- c("top_mast", "top_edge", "top_edge_dr", "top_mast_batch")

```


### top genes
```{r}
lapply(tops, head)
```
### log FC and pvalues summaries
```{r}
logfc <- lapply(names(tops), function(name){
  cat("\nMETHOD:", name, "\n")
  cat("\nlog2FC\n")
  print(summary(tops[[name]]$logFC))
  print(summary(tops[[name]]$avg_log2FC)
        )
  }
  )
pval <- lapply(names(tops), function(name){
  cat("\nMETHOD:", name, "\n")
  cat("\nPvalue adjusted\n")
  print(summary(tops[[name]]$FDR))
  print(summary(tops[[name]]$p_val_adj)
        )
  }
  )
```

## Comparisons gene names

```{r}
# extract just the gene names, and add helpful names
# b = batch correct
# dr = detection rate correct

 mast_dr <- row.names(top_mast)
 mast_dr_b <- row.names(top_mast_batch)
 edgeR_b <- row.names(top_edge)
 edgeR_dr_b <- row.names(top_edge_dr)
```
 Print for each comparison the number of geens in common, and which ones they are. 
 
```{r}
compare<- function(method1, method2, genes = TRUE){
  print(sum(method1 %in% method2))
  if(genes == TRUE){
print(method1[method1 %in% method2])
  }
}
```

#### The effect of adding detection rate to edgeR:
```{r}
# compare the edgeR between them
compare(edgeR_b,edgeR_dr_b)
```
Only 6 genes in common


#### The effect of adding batch correction to MAST:
```{r}
# compare the edgeR between them
compare(mast_dr,mast_dr_b, genes=FALSE)
```
MASTs compared between them
 
```{r}
sum(mast_dr %in% mast_dr_b)
identical(mast_dr, mast_dr_b)
identical(sort(mast_dr), sort(mast_dr_b))
```
 
Almost identical, the same sets of genes are selected, only the values change
 
#### MAST vs edgeR

without detection rate
```{r}
# compare the ones with dr between them
compare(edgeR_b, mast_dr_b)
```
with detectino rate

```{r}
# compare the ones with dr between them
compare(edgeR_dr_b, mast_dr_b)

```
